The Unicode Cookbook \linebreak for Linguists

### 2016-08-18 book/chapters/orthography_profiles.tex
minor changes to chapter

Our central aims for linguistics, to be approached with a Unicode-based
solution, are: (i) to improve the consistency of the encoding of sources, (ii)
to transparently document knowledge about the writing system (including
transliteration), and (iii) to do all of that in a way that is easy and quick to
manage for many different sources with many different writing systems. The
central concept in our proposal is the \textsc{orthography profile}, a simple
tab-separated CSV text file, that characterizes and documents a writing system.
We also offer basic implementations in Python and R to assist with the
production of such files, and to apply orthography profiles for consistency
testing, grapheme tokenization and transliteration. Not only can orthography
profiles be helpful in the daily practice of linguistics, they also succinctly
document the orthographic details of a specific source, and, as such, might
fruitfully be published alongside sources (e.g.~in digital archives). Also, in
high-level linguistic analyzes in which the graphemic detail is of central
importance (e.g.~phonotactic or comparative-historical studies), orthography
profiles can transparently document the decisions that have been taken in the
interpretation of the orthography in the sources used.

Given these goals, Unicode locale descriptions (see Section~\ref{terminology})
might seem like the ideal orthography profiles. However, there are various
practical obstacles preventing the use of such locale descriptions in the daily
linguistic practice, namely: (i) the XML-structure is too verbose to easily and
quickly produce or correct manually, (ii) locale descriptions are designed for a
wide scope on information (like date formats or names of weekdays) most of which
is not applicable for documenting writing systems, and (iii) most crucially,
even if someone made the effort to produce a technically correct locale
description for a specific source at hand, then it is nigh impossible to deploy
the description. This is because a locale description has to be submitted to and
accepted by the Unicode Common Locale Data Repository. The repository is
(rightly so) not interested in descriptions that only apply to a limited set of
sources (e.g.~descriptions for only a single dictionary).

@@

Finally, to arrive at the graphemic tokenization in
(3), tailored grapheme clusters are needed (as for example specified in an
orthography profile). For example, this orthography profile would specify that
the sequence of characters, and form a single grapheme, and that and form a
grapheme. The orthography profile could also specify orthographic rules,
e.g.~when tokenization graphemes, in say English, the in the forms and should be
treated as distinct sequences depending on their contexts.

@@

An orthography profile describes the Unicode code points, characters, graphemes
and orthographic rules in a writing system. An orthography profile is a
language-specific (and often even resource-specific) description of the units
and rules that are needed to adequately model a writing system. An important
assumption of our work is that we assume a resource is encoded in Unicode (or
has been converted to Unicode). Any data source that the Unicode Standard is
unable to capture, will also not be captured by an orthography profile.

@@

The logic of orthographic parsing ...

Note that to deal with ambiguous parsing cases, we can use the Unicode approach
using the zero width joiner. This is actually a non-joiner (the name is
confusing): the idea is to add this character into the text to identify cases in
which a sequence of characters is not supposed to be a complex grapheme (even
though the sequence is in the orthography profile)

>>>>

Our central aims for linguistics, to be approached with a Unicode-based
solution, are: (i) to improve the consistency of the encoding of sources, (ii)
to transparently document knowledge about the writing system (including
transliteration), and (iii) to do all of that in a way that is easy and quick to
manage for many different sources with many different writing systems. The
central concept in our proposal is the \textsc{orthography profile}, a simple
tab-separated text file, that characterizes and documents a writing system.
We also offer basic implementations in Python and R to assist with the
production of such files, and to apply orthography profiles for consistency
testing, grapheme tokenization and transliteration. Not only can orthography
profiles be helpful in the daily practice of linguistics, they also succinctly
document the orthographic details of a specific source, and, as such, might
fruitfully be published alongside sources (e.g.~in digital archives). Also, in
high-level linguistic analyzes in which the graphemic detail is of central
importance (e.g.~phonotactic or comparative-historical studies), orthography
profiles can transparently document the decisions that have been taken in the
interpretation of the orthography in the sources used.

Given these goals, Unicode locale descriptions (see
Section~\ref{the-unicode-approach}) might seem like the ideal orthography
profiles. However, there are various practical obstacles preventing the use of
such locale descriptions in the daily linguistic practice, namely: (i) the
XML-structure is too verbose to easily and quickly produce or correct manually,
(ii) locale descriptions are designed for a wide scope on information (like date
formats or names of weekdays) most of which is not applicable for documenting
writing systems, and (iii) most crucially, even if someone made the effort to
produce a technically correct locale description for a specific source at hand,
then it is nigh impossible to deploy the description. This is because a locale
description has to be submitted to and accepted by the Unicode Common Locale
Data Repository. The repository is (rightly so) not interested in descriptions
that only apply to a limited set of sources (e.g.~descriptions for only a single
dictionary).

@@

Finally, to arrive at the graphemic tokenization in (3), tailored grapheme
clusters are needed (possibly specified in an orthography profile). For example,
an orthography profile might specify that the sequence of characters <tsʰ> form
a single grapheme. The orthography profile could also specify orthographic
rules, e.g.~when tokenizing graphemes, in say English, the sequences <sh> in the
forms <mishap> and <mishmash> should be treated as distinct sequences depending
on their contexts.

@@

An orthography profile describes the Unicode code points, characters, graphemes
and orthographic rules in a writing system. An orthography profile is a
language-specific (and often even resource-specific) description of the units
and rules that are needed to adequately model a writing system. An important
assumption is that we assume a resource is encoded in Unicode (or
has been converted to Unicode). Any data source that the Unicode Standard is
unable to capture, will also not be captured by an orthography profile.

@@

The crucial difference between our current proposal and traditional
computational approaches to transliteration is the strict separation between
tokenization and transliteration. Most computational approaches to
transliteration are based on finite-state transducers (including the
transliteration as described in the Unicode Locale Data Markup
Language.\footnote{See \url{http://www.unicode.org/reports/tr35/}} Finite-state 
transducers attempt to describe the mapping from input to output string directly 
as a set of rewrite rules. Although such systems are computationally well 
understood, we feel that they are not well-suited for day-to-day linguistic 
practice. First, by forcing a first step of grapheme tokenization, our system 
tries to keep close to the logic of the writing system. Second, by separating 
tokenization from transliteration there is no problem with `feeding' and 
`bleeding' of rules, common with transducers (cf.~Section~\ref{rule-ordering}
).

Note that to deal with ambiguous parsing cases, it is still possible to use the
Unicode approach including the \textsc{zero-width non-joiner} character at
\uni{200C} into the text. The idea is to add this character into the text to
identify cases in which a sequence of characters is \textit{not} supposed to be
a complex grapheme cluster (even though the sequence is in the orthography
profile).
